---
title: "Exploring the Simpsons Dataset"
subtitle: "INFO 526 - Summer 2026 - Final Project"
author: 
  - name: "Daniel Spangler"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "Using the Simpsons dataset to analyze episode rating and character speaking lines over seasons"
format:
   html:
    code-tools: true
    code-overflow: wrap
    embed-resources: true
    echo: false
editor: visual
execute:
  warning: false
  echo: false
---

## Abstract

This project provides visualizations to determine how IMDB ratings and character speaking lines have changed across seasons or eras of the Simpsons. This will be done by showing a box plot of the different IMDB ratings for each show runner era, showing the median rating and ranges of outcomes. The analysis will also include a heat map showing the top 10 characters who speak the most lines per era.

## Introduction

The data set comes from Tidy Tuesday 2025-02-04 and is all about the Simpsons, which is a very popular adult animated sitcom. This data was originally from The Simpsons Dataset on Kaggle. The data set contains four tables each highlighting a different aspect of the show. The tables include characters, episodes, locations, and script lines. The data is very large (with the show having so many seasons), with the character data frame including 6722 rows and script lines data frame containing 31793 rows.

Link to data set: <https://www.kaggle.com/datasets/prashant111/the-simpsons-dataset>

```{r}
#| label: loading necessary packages
#| message: false

library(tidyverse)
library(tidytuesdayR)

```

```{r}
#| label: load data set
#| message: false

# loading the data locally
simpsons_characters   <- read_csv("data/simpsons_characters.csv") # characters dataset
simpsons_episodes     <- read_csv("data/simpsons_episodes.csv") # episodes dataset
simpsons_locations    <- read_csv("data/simpsons_locations.csv") # locations dataset
simpsons_script_lines <- read_csv("data/simpsons_script_lines.csv") # script lines dataset
```

The data set is split into 4 different tables, including characters, episodes, locations, and script lines. The characters data set includes every character in the history of the Simpsons, with 6,722 rows and 4 columns including the character name and ID. The episodes data set contains one row per episode of the series over it's entire history, with values like IMDB ratings, the amount of viewers, the seasons, and release date. The episodes data set has 600 rows and 14 columns. The locations data set features all the unique locations in the Simpsons, with identifiers for each. There are 4,459 rows and 3 columns in this data set. Finally, the script lines data set is the largest, totaling with 158,271 and 13 columns. This shows every script line that has ever been said, including the specific season, character, and more.

The project will address two different questions:

## Questions

1.  How have the IMDB ratings of the Simpsons episodes changed over time?
2.  Which Simpsons characters have the most speaking lines during each era of the show, and how does this change over time?

## Question 1: How have the IMDB ratings of the Simpsons episodes changed over time?

### Introduction

The Simpsons TV show has been running for multiple decades at this point, first airing in 1989. Audiences have always argued that the quality of the show has declined since it first aired, but there is no real reasoning to this argument. What my question hopes to do is put real data behind this question, and analyze each show runner and the median IMDB ratings for their episodes to see if the quality of the show has actually declined, or it is just nostalgia talking. To do this, I will be using the episodes data set, with important columns like IMDB rating, season, and original air date. With this data set, I will have to create an era variable, to signify the four unique eras of the show. Seasons 1-8 will be known as the Classic Era, with these seasons having multiple different teams of show runners. Seasons 9-12 will be labeled as Scully, named after the show runner Mike Scully during that time. Seasons 13-19 will be called Jean SD, named after show runner Al Jean and how the show was in Standard Definition during that time. Lastly, any episodes in seasons 20+ will be named Jean HD, after the same show runner Al Jean, but now the show is in High Definition instead of SD like before.

I am interested in this question because I've heard the sentiment online about the more recent seasons being worse, and I wanted to see if the data could actually back it up. Everyone enjoys watching TV, and I think the Simpsons itself is really intriguing because it has been going on for so long, and so much has happened in the world during the time it has aired. That gives us great and accessible information we can do lots of studies on.

### Approach

To answer this question, I will be building two different types of visualizations. The first one that I will build is a box plot of of IMDB ratings by era of the Simpsons. These eras are defined above as Classic, Scully, Jean SD, and Jean HD. To do this, I must split the data into different era categories, then plot the data. This box plot is going to be meaningful for the audience to view because it doesn't shrink the topic down to one number like a mean would. It shows the median, and potential ranges for each era, highlighting not only better ratings, but the best and worst episodes each era.

The second graph that I will make is a scatter plot of IMDB ratings versus season with a trend line. Plotting the seasons on the x-axis instead of the era like before will give us a better view of the ratings on a season by season basis, instead of just the 4 eras like before. Although there are a lot more seasons, it won't be too much for the eye of the audience like the box plot of each season would. Once making the scatter plot, I will plot a trend line to see if the average rating really does go down as time moves on.

### Analysis

```{r}
#| label: create era variable
#| message: false

# creating era variable in episode data set
simpsons_episodes <- simpsons_episodes |>
  mutate(
    era = case_when(
      season >=1 & season <= 8 ~ "Classic (1-8)",
      season >= 9 & season <= 12 ~ "Scully (9-12)",
      season >= 13 & season <=19 ~ "Jean SD (13-19)",
      season >= 20 ~ "Jean HD (20+)"
    ),
    era = factor(era,levels = c("Classic (1-8)","Scully (9-12)","Jean SD (13-19)","Jean HD (20+)"))
  )
```

```{r}
#| label: create box plot
#| message: false

ggplot(simpsons_episodes, aes(x = era, y = imdb_rating, fill = era)) + 
  geom_boxplot(show.legend = FALSE) + # create box plot
  scale_fill_brewer(palette = "Set2") + # set color scheme
  labs(
    x = NULL,
    y = "IMDB Rating",
    title = "IMDB Ratings by Simpsons Era"
  ) + theme_minimal()
```

```{r}
#| label: create scatter plot
#| message: false

ggplot(simpsons_episodes, aes(x = season, y = imdb_rating)) + 
  geom_point(aes(color = era)) + # create scatter plot
  geom_smooth(method = "lm", color = "black", se = FALSE) + # plot straight trend line
  scale_x_continuous(breaks = seq(0, 35, by = 5)) + 
  scale_color_brewer(name = "Era", palette = "Set2") + # set same color scheme
  labs(
    x = "Season",
    y = "IMDB Rating",
    title = "IMDB Ratings across Seasons of the Simpsons"
  ) + theme_minimal()
```

### Discussion

The box plot in the first visualization suggests that as time has gone on, IMDB ratings actually have gotten worse and worse, supporting the argument of audiences saying that the show has gotten worse. You can see in the plot that in the Classic era, or seasons 1-8, the median IMDB rating falls at about an 8.25, and that is the highest median IMDB rating for all eras. In the Scully era this value falls to around 7.5, in the Jean SD it is about a 7, and finally in Jean HD the value is at a mere 6.75. This fully supports audience claims about the newer seasons being worse than the original, and it isn't just nostalgia talking. One interesting thing you can see from the plot is the range of outcomes for each era, as the Jean HD era shows episodes that are both higher rated, and lower rated than Jean SD, meaning they have a less predictable range of outcomes with some episodes being really good and others really bad.

The scatter plot reinforces the idea from the box plot, showing that as time goes on, the IMDB ratings for the Simpsons episodes get worse on average. You can see this explicitly by the trend line gained from doing the linear method on the data. In this trend line, you can very easily see that there is a negative slope of the line, as seasons increases, the IMDB rating on average decreases. This plot is also useful to see the ranges of outcomes for each season, similar to what we talked about before. You can see the explicit episodes that got high and low ratings for each season.

There are many reasons people speculate that Simpsons episodes have gotten worse over time, but this just tends to happen to all TV shows that go on for that long. As time goes on, they run out of story lines, new people are brought onto the show, and overall the audience just starts to lose interest. This has also been the case for other long lasting TV shows like Spongebob, Family Guy, and even the Office.

## Question 2: Which Simpsons characters have the most speaking lines during each era of the show, and how does this change over time?

### Introduction

Question 1 shows how the show was perceived by the audience through IMDB ratings, but question 2 highlights how those specific seasons were built, with what specific characters shining the brightest and having the most speaking lines. Question 2 will focus on the script lines data set which shows all the dialogue and lines ever spoken throughout all of the seasons of the Simpsons. The data set includes things like what the line was, which character spoke it, and what season is was during. Doing this analysis will help us understand not only which characters are the most popular during every season, but after relating it to the first question and first two visualizations we can see which characters are loved by the audience and make for good story lines, relating the seasons or eras that are the most popular with those specific characters.

I am interested in this question because online there have been a lot of conversations about which characters have taken a back seat in the more recent seasons. Allegedly, they have sidelined a lot of the minimal side characters and focused on Homer and the handful of more popular characters instead of diving into story lines all over the place like they did originally. I want to see if this is actually true, and if the amount of dialogue that Homer and the more popular characters have had has decreased since the first couple of seasons.

### Approach

To do this, the first thing I will do is determine the top 10 characters with the most lines throughout all the seasons of the show. These will be the default characters used in my analysis, and what we will focus on for every season. Instead of just doing graphs with the amount of lines, i decided to do them based on the percentage of lines that the character had in the certain season or era. After finding the top 10 characters with the most speaking lines, I determined the percentage of dialogue lines they spoke based on the season or era.

The first visualization will be a stacked bar chart with season on the x-axis and percent line count on the y-axis with the graph being separated into different colors for each character. The second visualization is going to be a heat map with the era on the x-axis and the top 10 speaking characters of all-time on the y-axis. The color or heat of the squares will be the percentage of dialogue they have during that specific era. A heat map will be good for this analysis because it will be a simple way to visualize the amount of character dialogue. It will be a bit easier to visualize with eras on the x-axis instead of seasons with the stacked bar chart, as that is a lot for the audience to take in.

### Analysis

```{r}
#| label: creating top 10
#| message: false

lines_by_season <- simpsons_script_lines |>
  left_join(
    simpsons_episodes |> select(id, season, era),
    by = c("episode_id" = "id")
  ) |>
  filter(!is.na(season), !is.na(raw_character_text))

top10_characters <- lines_by_season |>
  count(raw_character_text, sort = TRUE) |>
  slice_head(n = 10) |>
  pull(raw_character_text)

season_percent <- lines_by_season |>
  count(season, raw_character_text, name = "line_amount") |>
  group_by(season) |>
  mutate(pct_lines = line_amount / sum(line_amount) * 100) |>
  ungroup() |>
  filter(raw_character_text %in% top10_characters) |>
  mutate(raw_character_text = factor(raw_character_text, levels = rev(top10_characters)))

era_percent <- lines_by_season |>
  count(era, raw_character_text, name = "line_amount") |>
  group_by(era) |>
  mutate(pct_lines = line_amount / sum(line_amount) * 100) |>
  ungroup() |>
  filter(raw_character_text %in% top10_characters) |>
  mutate(raw_character_text = factor(raw_character_text, levels = rev(top10_characters)))

```

```{r}
#| label: stacked bar chart
#| message: false

ggplot(season_percent, aes(x = season, y = pct_lines, fill = raw_character_text)) + 
  geom_col() + 
  scale_x_continuous(breaks = seq(0, 35, by = 5)) + 
  scale_y_continuous(labels = scales::label_percent(scale = 1)) + 
  scale_fill_brewer(name = "Character", palette = "Set3") +
  labs(
    x = "Season",
    y = "Line Percentage",
    title = "Simpsons Top 10 Characters by Line Amounts per Season"
  ) + theme_minimal()
  
```

```{r}
#| label: stacked bar chart
#| message: false

ggplot(era_percent, aes(x = era, y = raw_character_text, fill = pct_lines)) + 
  geom_tile() + 
  scale_fill_distiller(name = "Line Percentage", palette = "Greens", direction = 1, labels = scales::label_percent(scale = 1)) + 
  geom_text(
    aes(label = paste0(round(pct_lines, 1), "%"),
        color = pct_lines > max(pct_lines) * 0.6),
    show.legend = FALSE
  ) + scale_color_manual(values = c("black", "white")) + 
  labs(
    x = "Era",
    y = NULL,
    title = "Heat Map of Character Dialogue by Simpsons Era"
  ) + theme_minimal()
  
```

### Discussion

The stacked bar chart in the first visualization turned out a little bit confusing for the eye, but overall what you can pull from it is that Homer has always been the main character of the show regardless of the season. You can see that he is in the purple on the bottom of the graph, and always has the most dialogue percentage. The other characters that make up a good chunk are Marge, Bart, and Lisa. After those characters, the percentage of dialogue really falls off, with side characters only accounting for around 10% of the season dialogue on average. Based on this graph, it is hard to see the exact change in percentage of character lines over the season, but the main takeaway is that the core 4 take up a large amount of the dialogue regardless of the season or era.

The heat map helps tell a bit more of the story, with exact values showing the falloffs and and rises of some certain characters. The first takeaway is the decline in percentage of dialogue from Bart Simpson. After having the second highest percentage of dialogue behind Homer in the Classic era with 11.8%, Bart never once reaches 10% again in the eras to follow. After the first era, he is overtaken by Marge, who has the second most lines in the series all-time. Additionally, There is a large decline in line percentage for Mr. Burns, who is seen as the main antagonist of the show. In the Classic era, he has a line percentage of 3.4% whereas in the following seasons he barely reaches over 2%. This one percent drop might not seem like a lot, but over the entire era with multiple seasons of shows, it shows something significant. All other characters are somewhat constant throughout the seasons, except for Moe Szyslak, who increases constantly from 1.4% to 2% to 2.2% then to 2.6% in the most recent era.

Going back to question 1, and analyzing the decreased IMDB rating over the years, there are a few ways to relate it to the second question. There are a bunch more outside factors, but specifically just looking at this graph, we could conclude that Bart and Mr. Burns are popular characters with good story lines. When these characters were decreased in the eras following the Classic era, the IMDB ratings also continued to decrease. This is a large assumption, but is interesting to think about in the context of the problem. To make this claim certain, there would have to be other studies that deals with audience popularity scores and surveys, but it is something that was interesting to pull from this visualization.
