---
title: "Exploring the Simpsons Dataset"
subtitle: "Final Project Proposal"
author: 
  - name: "Daniel Spangler"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "Using the Simpsons dataset to analyze episode rating and character appearances over seasons"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false

library(tidyverse)
library(tidytuesdayR)
```

## Dataset

A brief description of your dataset including its provenance, dimensions, etc. as well as the reason why you chose this dataset.

Make sure to load the data and use inline code for some of this information.

```{r}
#| label: load-dataset
#| message: false

# loading the data locally
simpsons_characters   <- read_csv("data/simpsons_characters.csv") # characters dataset
simpsons_episodes     <- read_csv("data/simpsons_episodes.csv") # episodes dataset
simpsons_locations    <- read_csv("data/simpsons_locations.csv") # locations dataset
simpsons_script_lines <- read_csv("data/simpsons_script_lines.csv") # script lines dataset
```

```{r}
#| label: dataset-dims

# characters dataset has 6722 rows and 4 columns
dim(simpsons_characters)

# episodes dataset has 151 rows and 14 columns
dim(simpsons_episodes)

# locations dataset has 4459 rows and 3 columns
dim(simpsons_locations)

# script lines dataset has 31,793 rows and 13 columns
dim(simpsons_script_lines)
```

The data set that I chose comes from Tidy Tuesday 2025-02-04 and is all about the Simpsons, which is a very popular adult animated sitcom. This data was originally from The Simpsons Dataset on Kaggle. The data set contains four tables each highlighting a different aspect of the show. The tables include characters, episodes, locations, and script lines. The data is very large (with the show having so many seasons), with the character data frame including 6722 rows and script lines data frame containing 31793 rows.

I chose this data set because it is really interesting to look at the success of each different season, and see if that can be attributed to a certain character or location. Any analysis that I can do on movies or TV I always find super interesting, and because of the immense history and quality of this show, I jumped at the chance to do a project on it.

Link to original data set: <https://www.kaggle.com/datasets/prashant111/the-simpsons-dataset>

## Questions

1.  How have the IMDB ratings of the Simpsons episodes changed over time?
2.  Which Simpsons characters have the most speaking lines each season, and how does this change over time?

## Analysis plan

Question 1

Variables:

-   season, imdb_rating, us_viewers_in_millions, original_air_year

-   Create categorical variable for era

    -   Classic Era (Seasons 1-8)

    -   Scully Era (Seasons 9-12)

        -   Named after showrunner Mike Scully

    -   Jean SD Era (Seasons 13-19)

        -   Named after Al Jean

    -   Jean HD Era (Seasons 20-Today)

    -   Sourced from <https://nohomers.net/forums/index.php?threads/the-simpsons-eras-division.43685/>

Planned visualizations:

-   Boxplot of average imdb_rating per era

-   Bar chart of average imdb_rating per era, with trend lines

Question 2

Variables:

-   raw_character_text, episode id from script lines data set

-   season from episodes data set

Planned visualizations:

-   Stacked bar chart showing top 10 characters with most lines

-   Heatmap (also top 10 characters)

## Weekly Plan of Attack

Week 1: Data Exploration, Setup, and Proposal

Set up the Github repository and brainstorm what data set/questions I would like to explore. After deciding, I can start the proposal and start brainstorming how I would like to answer these questions, using what different visualizations. I will import the four data sets into the proposal, and see the dimensions of each.

Week 2: Continue Proposal

I can now fill out the Analysis Plan section of the proposal, giving specific information of what visualizations I am going to create, and what variables are needed to complete this task. I will define the different 'eras' that I mentioned in question 1, and give research as to why I chose those values.

Week 3: Peer Review

I will perform peer reviews on two projects from other students. Using this information and seeing what they wrote in their proposal, I can incorporate anything that I believe I missed. Also, looking at their reviews of my project, I can add more or edit my project to improve it.

Week 4: Revise proposal + begin write-up (`index.qmd`)

Using the Issues from our teacher, I can revise my proposal and add anything that I missed from the template. Once I did this, I will begin my write up for the project by bringing in the data, finalizing the eras that I mentioned before, and correctly joining data sets to merge characters and script lines for question 2. I will complete a rough draft of what my visualizations will look like.

Week 5: Work on project

I will continue working on my project, improving all of the plots and writing in the reflection sections of each plot, giving the audience a look into why these plots are important, and the reasoning I chose them.

Week 6: Finalize all materials and record presentation

I will finalize the presentation and record it, while also making sure everything is in the correct place for the project. I will double check that the website is working, and correctly displays all the information I was hoping it would.
