---
title: "Exploring the Simpsons Dataset"
subtitle: "Final Project Proposal"
author: 
  - name: "Daniel Spangler"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "Using the Simpsons dataset to analyze episode rating and character appearances over seasons"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false

library(tidyverse)
library(tidytuesdayR)
```

## Dataset

A brief description of your dataset including its provenance, dimensions, etc. as well as the reason why you chose this dataset.

Make sure to load the data and use inline code for some of this information.

```{r}
#| label: load-dataset
#| message: false

# loading the data locally
simpsons_characters   <- read_csv("data/simpsons_characters.csv") # characters dataset
simpsons_episodes     <- read_csv("data/simpsons_episodes.csv") # episodes dataset
simpsons_locations    <- read_csv("data/simpsons_locations.csv") # locations dataset
simpsons_script_lines <- read_csv("data/simpsons_script_lines.csv") # script lines dataset
```

```{r}
#| label: dataset-dims

# characters dataset has 6722 rows and 4 columns
dim(simpsons_characters)

# episodes dataset has 151 rows and 14 columns
dim(simpsons_episodes)

# locations dataset has 4459 rows and 3 columns
dim(simpsons_locations)

# script lines dataset has 31,793 rows and 13 columns
dim(simpsons_script_lines)
```

The data set that I chose comes from Tidy Tuesday 2025-02-04 and is all about the Simpsons, which is a very popular adult animated sitcom. This data was originally from The Simpsons Dataset on Kaggle. The data set contains four tables each highlighting a different aspect of the show. The tables include characters, episodes, locations, and script lines. The data is very large (with the show having so many seasons), with the character data frame including 6722 rows and script lines data frame containing 31793 rows.

I chose this data set because it is really interesting to look at the success of each different season, and see if that can be attributed to a certain character or location. Any analysis that I can do on movies or TV I always find super interesting, and because of the immense history and quality of this show, I jumped at the chance to do a project on it.

Link to original data set: <https://www.kaggle.com/datasets/prashant111/the-simpsons-dataset>

## Questions

1.  How have the IMDB ratings of the Simpsons episodes changed over time?
2.  Which Simpsons characters have the most speaking lines each season, and how does this change over time?

## Analysis plan

Question 1

Variables:

-   season, imdb_rating, us_viewers_in_millions, original_air_year

-   Create categorical variable for era

    -   Classic Era (Seasons 1-8)

    -   Scully Era (Seasons 9-12)

        -   Named after showrunner Mike Scully

    -   Jean SD Era (Seasons 13-19)

        -   Named after Al Jean

    -   Jean HD Era (Seasons 20-Today)

    -   Sourced from <https://nohomers.net/forums/index.php?threads/the-simpsons-eras-division.43685/>

Planned visualizations:

-   Boxplot of average imdb_rating per era

-   Bar chart of average imdb_rating per era, with trend lines

Question 2

Variables:

-   raw_character_text, episode id from script lines data set

-   season from episodes data set

Planned visualizations:

-   Stacked bar chart showing top 10 characters with most lines

-   Heatmap (also top 10 characters)

## Weekly Plan of Attack

Week 1: Data Ingestion & Environmental Setup

Set up repository and working environment, choose data set, and ingest the dataset via the GitHub URL. Run initial glimpse(), dim(), and summary() checks. Identify missing values (NA values), specifically in the Release Date or Run Time columns, and decide how to handle them (e.g., filtering them out or looking up missing pieces).

Week 2: Feature Engineering & Data Cleaning

Create categorical era buckets and content length buckets. Merge data and flag data into "Movies" vs. "TV Shows". Calculate the content age/decay metrics. Print out small table summaries using count() to make sure custom categories have a healthy balance of rows.

Week 3: Exploratory Data Analysis (EDA) & Summary Statistics

Calculate standard numeric summaries (mean, median, standard deviation) of Views and Hours Viewed across different categorical buckets. Look at the skewness of viewership data. Because streaming hits have massive outliers, decide if it is necessary to apply a log transformation to numeric axes for upcoming charts.

Week 4: Data Visualization (Drafting the Core Plots)

Develop Question 1 plots: The scatter plot tracking popularity decay over time, and the box plots comparing release eras. Develop Question 2 plots: The faceted scatter plots breaking down runtime trends by Movie vs. Show. Apply clean axis labels, custom scales, and clear legends.

Week 5: Statistical Interpretation & Narrative Writing

Draft the written sections in Quarto document beneath code chunks. Interpret the charts: Does an older era hold a surprising amount of nostalgia/lasting popularity? Is there an optimal movie runtime that maximizes views? Use print statements within code to dynamically display key takeaway numbers directly inside report text.

Week 6: Quarto Polishing, Code Review, and Formatting

Render and double check all content. Finalize repository and record presentation.
