---
title: "Exploring the Simpsons Dataset"
subtitle: "Final Project Proposal"
author: 
  - name: "Daniel Spangler"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "Using the Simpsons dataset to analyze episode rating and character appearances over seasons"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false
library(tidyverse)
library(tidytuesdayR)
```

## Dataset

```{r}
#| label: load-dataset
#| message: false
```

A brief description of your dataset including its provenance, dimensions, etc. as well as the reason why you chose this dataset.

Make sure to load the data and use inline code for some of this information.

```{r}
tuesdata <- tidytuesdayR::tt_load(2025, week = 5)

simpsons_characters <- tuesdata$simpsons_characters
simpsons_episodes <- tuesdata$simpsons_episodes
simpsons_locations <- tuesdata$simpsons_locations
simpsons_script_lines <- tuesdata$simpsons_script_lines
```

```{r}
dim(simpsons_characters)
dim(simpsons_episodes)
dim(simpsons_locations)
dim(simpsons_script_lines)
```

The data set that I chose comes from Tidy Tuesday 2025-02-04 and is all about the Simpsons, which is a very popular adult animated sitcom. This data was originally from The Simpsons Dataset on Kaggle. The data set contains four tables each highlighting a different aspect of the show. The tables include characters, episodes, locations, and script lines. The data is very large (with the show having so many seasons), with the character data frame including 6722 rows and script lines data frame containing 31793 rows.

I chose this data set because it is really interesting to look at the success of each different season, and see if that can be attributed to a certain character or location. Any analysis that I can do on movies or TV I always find super interesting, and because of the immense history and quality of this show, I jumped at the chance to do a project on it.

Link to original data set: <https://www.kaggle.com/datasets/prashant111/the-simpsons-dataset>

## Questions

The two questions you want to answer.

## Analysis plan

-   A plan for answering each of the questions including the variables involved, variables to be created (if any), external data to be merged in (if any).
